Compositional Certified Resource Bounds\ifshort{\vspace{-.15cm}}{}

### 2014-07-04 paper/main.tex
more updates to the intro

For functional programs, there exist resource analysis systems---based
on type-based amortized resource analysis~\cite{Jost03,
  HoffmannAH12}---that can automatically derive complex polynomial
bounds with tight constant factors.  This type-based amortized
resource analysis has been an inspiration in the design of the
quantitative Hoare logic and is therefore a natural candidate to
automate the reasoning.  However, it is a well-known problem that
type-based amortized resource analysis works well for functional
programs with pattern matching but not for C-like programs with
control flow that depends integer arithmetic.  For example, it is a
long-time open problem how to use type-based amortized analysis to
derive bounds for functions whose resource consumption depends on
(possibly negative) integers and non-sequential control flow as
introduced by \code{break} and \code{return} statements.
Unfortunately, these are the functions \iffull{with non-constant resource
consumption}{} that are most common in CertiKOS and other system
software.

Our \emph{main contribution} is an automatic amortized resource
analysis for CompCert Clight that computes derivations in the
quantitative Hoare logic.  It is the first automatic amortized
resource analysis that can derive bounds that depend on negative
integers. It also handles programs with mutually recursive functions as
well as \code{break} and \code{return} statements.  Using the
potential-based approach of amortized analysis allows us to
use a single mechanism to analyze programs that need special treatment in other
techniques.  For instance, we can naturally handle
\itemskip
\begin{itemize}
\item interaction between \emph{sequential loops or function calls} through
  size changes of variables,
\itemskipIn
\item \emph{nested loops} that influence each others number of iterations,
\itemskipIn
\item and \emph{tricky iteration patterns} as found for instance in
  the Knuth-Morris-Pratt algorithm for string search.
\end{itemize}
\itemskip
%
The main innovation that makes the analysis practical for C is the use
of interval sizes in potential functions instead of the sizes of
variables.  This leads to precise bounds of programs whose resource
consumption can be described as a function of sizes $|[x,y]| =
\max(0,y-x)$ of intervals of integer variables.  Such bounds arise for
instance from standard \emph{for loops} \code{for (i = x; i+K \leq y;
  i=i+K)} where the step-size $K>0$ is a constant.
%

@@

We have evaluated the automatic analysis with system code and examples
from the literature.  \pref{app:cat} contains more than $30$
challenging loop and recursion patterns that we collected from open
source software and the literature.  Our analysis can find
asymptotically tight bounds for $30$ all but 3 of these patterns and
in most cases the derived constant factors are tight.  To compare our
automatic analyzer with existing techniques, we tested our examples
with the tools KoAT~\cite{BrockschmidtEFFG14},
PUBS~\cite{AlbertAGPZ12}, Rank~\cite{AliasDFG10}, and
LOOPUS~\cite{SinnZV14}.  We also contacted the authors of
SPEED~\cite{GulwaniMC09} but were able to obtain a tool to perform a
comparison.  Our experiments show that the bounds that we derive are
often more precise than those derived by existing tools.  Only
LOOPUS~\cite{SinnZV14}, which also uses amortization techniques, is
able to achieve a similar precision.

Several micro benchmarks demonstrate the practicability and
expressiveness of the quantitative Hoare logic.  For example, we
derive a logarithmic bound for a binary search function, a bound that
depends on the contents of an array to describe the exact cost of a
function that finds the maximal element in an array, and a linear
bound that amortizes the cost of $n$ successive increments to a binary
counter.  XXX In a larger case study, we show how the quantitative
logic can be used to derive the amortized cost for data-structure
operations for a table that expands or shrinks depending on the load
factor.

>>>>

For functional programs, there exist resource analysis systems---based
on type-based amortized resource analysis~\cite{Jost03,
  HoffmannAH12}---that can automatically derive complex polynomial
bounds with tight constant factors.  This type-based amortized
resource analysis has been an inspiration in the design of the
quantitative Hoare logic and is therefore a natural candidate to
automate the reasoning.  Type-based amortized resource analysis works
well for functional programs with pattern matching but it has been a
long-time open problem to apply it to C-like programs with control
flow that depends integer arithmetic, negative numbers, and non-linear
control flow.
% example, it is a
% long-time open problem how to use type-based amortized analysis to
% derive bounds for functions whose resource consumption depends on
% (possibly negative) integers and non-sequential control flow as
% introduced by \code{break} and \code{return} statements.
% Unfortunately, these are the functions \iffull{with non-constant resource
% consumption}{} that are most common in CertiKOS and other system
% software.

Our \emph{main contribution} is an automatic amortized resource
analysis for CompCert Clight that computes derivations in the
quantitative Hoare logic.  It is the first automatic amortized
resource analysis that can derive bounds that depend on negative
integers. It also handles programs with mutually recursive functions
as well as \code{break} and \code{return} statements.  The
potential-based approach of amortized analysis provides a single
mechanism for analyzing programs that need special treatment in other
techniques; including
\itemskip
\begin{itemize}
\item interaction between \emph{sequential loops or function calls} through
  size changes of variables,
\itemskipIn
\item \emph{nested loops} that influence each others number of iterations,
\itemskipIn
\item and \emph{tricky iteration patterns} as found for instance in
  the Knuth-Morris-Pratt algorithm for string search.
\end{itemize}
\itemskip
%
The main innovation that makes the analysis practical for C is the use
of interval sizes in potential functions instead of the sizes of
variables.  This leads to precise bounds of programs whose resource
consumption can be described as a function of sizes $|[x,y]| =
\max(0,y-x)$ of intervals of integer variables.  Such bounds arise for
instance from standard \emph{for loops} \code{for (i = x; i+K \leq y;
  i=i+K)} where the step-size $K>0$ is a constant.
%

@@

We have evaluated the automatic analysis with system code and examples
from the literature.  \pref{app:cat} \ifshort{in the extended version
  of this article\cite{anon_extended}}{} contains more than $30$
challenging loop and recursion patterns that we collected from open
source software and the literature.  Our analysis can find
asymptotically tight bounds for all but 3 of these patterns and in
most cases the derived constant factors are tight.  To compare our
automatic analyzer with existing techniques, we tested our examples
with the tools such as KoAT~\cite{BrockschmidtEFFG14},
Rank~\cite{AliasDFG10}, and LOOPUS~\cite{SinnZV14}.  Our experiments
show that the bounds that we derive are often more precise than those
derived by existing tools.  Only LOOPUS~\cite{SinnZV14}, which also
uses amortization techniques, is able to achieve a similar precision.
%
Several micro benchmarks demonstrate the practicability and
expressiveness of the quantitative Hoare logic.  For example, we
derive a logarithmic bound for a binary search function, a bound that
depends on the contents of an array to describe the exact cost of a
function that finds the maximal element in an array, and a linear
bound that amortizes the cost of $n$ successive increments to a binary
counter.
